---
title: "Grain export"
output:
  html_document:
    anchor_sections: true
    highlight: pygments
    css: [css/base.css, css/downlit.css]
    md_extensions: -smart
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo=TRUE, warning=FALSE, message=FALSE,
  fig.align="center", fig.width=9.5, dev="svglite",
  class.source="sourceCode"
)

# Enable console highlighting, see `?fansi::set_knit_hooks`

old.hooks <- fansi::set_knit_hooks(knitr::knit_hooks)

styles <- c(
  getOption('fansi.style'),
  "PRE.fansi CODE {background-color: transparent;}",
  "PRE.fansi-error {background-color: transparent;}",
  "PRE.fansi-warning {background-color: transparent;}",
  "PRE.fansi-message {background-color: transparent;}"
)

old.hooks <- c(
  old.hooks,
  fansi::set_knit_hooks(
    knitr::knit_hooks,
    which=c('warning', 'error', 'message'),
    style=styles
  )
)

options(crayon.enabled = TRUE)
```

## Techniques employed

- Data wrangling:
  - Filtering data
  - Working with dates: months, years
  - Anti-joining data
  - Grouping and summarising data
  - Reordering factor levels by the values of a function of a secondary variable

- Data visualisation:
  - Line plots
  - Facetting by a secondary variable
  - Highlighting levels of interest within a facet

- String operations:
  - Case insensitive string matching
  - Partial string matching
  - Removing a substring


## About the data

The export of Canadian grains has been collected monthly between 1922 to 2018. This data set
includes information on the final destination as well as the type of grain being exported (e.g.
wheat, oats, barley, etc.).

More information on this record can be found
[here](https://open.canada.ca/data/en/dataset/33a5e534-b62a-45b7-b36c-ddc16d53f6a0).

The data and accompanying metadata can be downloaded together as a .zip file from
[here](https://www150.statcan.gc.ca/n1/tbl/csv/32100008-eng.zip).


## Packages

```{r}
library(tidyverse)
library(lubridate)
library(gghighlight)
theme_set(theme_bw())
```


## Obtaining the data

For these demos, I will be storing all data files in a folder called `data` with subfolders for
each topic.

```{r, eval=FALSE}
if (!dir.exists("./data/grains")) {
  dir.create("./data/grains", recursive=TRUE)
}

download.file(
  "https://www150.statcan.gc.ca/n1/tbl/csv/32100008-eng.zip",
  destfile = "./data/grains/grains.zip"
)

unzip(
  "./data/grains/grains.zip",
  exdir = "./data/grains"
)
```

```{r}
list.files("./data/grains", pattern="csv")
```


## Reading in the data

```{r}
grains <- read_csv("./data/grains/32100008.csv")

glimpse(grains)
```

The variables of interest are:

| Variable             | Description                            |
|----------------------|----------------------------------------|
| `REF_DATE`           | Year and month of record               |
| `Commodity`          | Type of grains exported                |
| `VALUE`              | Amount of grains exported in tonnes    |
| `Destinations`       | Destination where grains were exported |

```{r}
grains <- grains %>%
  select(REF_DATE, Commodity, VALUE, Destinations) %>%
  rename(
    date = REF_DATE,
    type = Commodity,
    value = VALUE,
    destination = Destinations
  )
```

## Dividing the data

From calling `glimpse()` it is clear that a portion of the data contains summaries.

```{r}
grains %>%
  distinct(destination) %>%
  pull(destination) %>%
  head(n=15)
```

Let's divide the data into two chunks: one that contains the summaries, and one without the
summaries. We can search for all summaries by looking for values of `destination` that contain the
term "total" (case-insensitive search). We can then isolate the remaining data by anti-joining
the original data set with the summaries, which will return all data that did not appear in the
summary data. Finally, we can drop any rows that contain missing values.

```{r}
grains_totals <- grains %>%
  filter(str_detect(destination, pattern="(?i)total")) %>%
  drop_na()

grains <- grains %>%
  anti_join(grains_totals, by="destination") %>%
  drop_na()
```


## Making proper dates

We can make proper dates of the `date` column by using `lubridate::ym()`. Let's also create a new
variable to keep track of the year of observations and move the date-related columns together.

```{r}
grains_totals <- grains_totals %>%
  mutate(
    date = ym(date),
    Year = year(date)
  ) %>%
  select(date, Year, type, value, destination)

grains <- grains %>%
  mutate(
    date = ym(date),
    Year = year(date)
  ) %>%
  select(date, Year, type, value, destination)
```


## Creating annual summaries

```{r}
grains_totals %>%
  group_by(destination) %>%
  slice_min(date) %>%
  ungroup() %>%
  distinct(date)

grains_totals %>%
  group_by(destination) %>%
  slice_max(date) %>%
  ungroup() %>%
  distinct(date)
```

It can be seen that the range of the dates in the data set is quite large. Since the data is also
occurring monthly, any visualisations that use the raw dates will appear too busy. Therefore, for
the following examples, let's only focus on the annual values. Of course, further analysis could be
performed at the monthly level or between a smaller range of dates.

```{r}
grains_totals_annual <- grains_totals %>%
  group_by(destination, type, Year) %>%
  summarise(value = sum(value), .groups="drop")

grains_annual <- grains %>%
  group_by(destination, type, Year) %>%
  summarise(value = sum(value), .groups="drop")
```


## Visualising total annual export to various regions, by grain type

Let's first make another partition for total annual export to all destinations. Instead of
filtering on the condition of:

```{r, eval=FALSE}
destination == "Total exports, all destinations"
```

we can use `str_detect()` to perform partial string matching and do a bit less typing.

```{r}
grains_totals_annual_all <- grains_totals_annual %>%
  filter(str_detect(destination, "all destinations"))
```

Plotting the data as is:

```{r}
ggplot(grains_totals_annual_all, aes(x=Year, y=value, colour=type))+
  geom_line(alpha=0.6)
```


### Improving the plot

- It is unclear what the y-axis is measuring
- Y-axis is in scientific notation which requires the reader to do math
- It may be better to facet by grain type, which will increase the amount of available plotting
space by getting rid of the legend
- It may also be nice to facet by grain type, while keeping all other grain types greyed out in each
facet as a reference
- The facetted plot can be made more convenient to read if the facetted grain types were ordered
by their respective all-time maximums


#### Avoiding scientific notation on the y-axis

To avoid scientific notation on the y-axis, we can scale the values by one million. Thus, the new
y-values will represent the amount exported in millions of tonnes.

```{r}
grains_totals_annual_all <- grains_totals_annual_all %>%
  mutate(value = value / 1e6)
```


#### Ordering the grain types by their all-time maximums

The `type` column is the variable whose values are to be reordered. These values will be reordered
by the maximum value of each grain type, in descending order.

```{r}
grains_totals_annual_all <- grains_totals_annual_all %>%
  mutate(type = fct_reorder(type, value, .fun=max, .desc=TRUE))
```


#### Building the final plot

```{r, fig.height=9}
ggplot(grains_totals_annual_all, aes(x=Year, y=value, colour=type))+
  geom_line(show.legend=FALSE)+
  gghighlight(use_direct_label=FALSE)+
  facet_wrap(~type)+
  labs(
    y="Amount exported (million tonnes)",
    title="Annual grain amounts exported to all destinations, by grain type",
    subtitle="1922 to 2018 (where data is available)"
  )
```


#### Additional plots

```{r}
grains_totals_annual %>%
  distinct(destination)
```

The above plotting procedure can be repeated for the remaining destination totals, as desired.


## Visualising total annual wheat export (excluding durum), by region

The plotting procedure is similar to the previous plot, only now, we shall filter the data for
grain type `"Wheat, excluding durum"` and facet by the destination (excluding
`"Total exports, all destinations"`). In addition, we should also remove the word `"Total"` from
the destination names that we will facet upon.


#### Filtering the data

Here, we wish to include only export data for grains of type `"Wheat, excluding durum"` and
destinations other than `"Total exports, all destinations"`, since the total exports to all
destinations was already seen in the previous plot.

```{r}
total_annual_wheat_export <- grains_totals_annual %>%
  filter(str_detect(type, "Wheat, excluding")) %>%
  filter(str_detect(destination, "all destinations", negate=TRUE))
```


#### Avoiding scientific notation on the y-axis

To avoid scientific notation on the y-axis, we can scale the values by one million. Thus, the new
y-values will represent the amount exported in millions of tonnes.

```{r}
total_annual_wheat_export <- total_annual_wheat_export %>%
  mutate(value = value / 1e6)
```


#### Adjusting the destination labels

The word `"Total"` (and the space that comes after it), are to be removed from the destination
labels.

```{r}
total_annual_wheat_export <- total_annual_wheat_export %>%
  mutate(destination = str_remove(destination, pattern="Total "))
```


#### Ordering the destinations by their all-time maximums

The `destination` column is the variable whose values are to be reordered. These values will be
reordered by the maximum value of wheat (excluding durum) exported, in descending order.

```{r}
total_annual_wheat_export <- total_annual_wheat_export %>%
  mutate(destination = fct_reorder(destination, value, .fun=max, .desc=TRUE))
```


#### Building the plot

```{r, fig.height=9}
ggplot(total_annual_wheat_export, aes(x=Year, y=value, colour=destination))+
  geom_line(show.legend=FALSE)+
  gghighlight(use_direct_label=FALSE)+
  facet_wrap(~destination)+
  labs(
    y="Amount exported (million tonnes)",
    title="Annual amounts of wheat exported, by destination",
    subtitle="1971 to 2018 (where data is available)"
  )
```


#### Additional plots

```{r}
grains_totals_annual %>%
  distinct(type)
```

The above plotting procedure can be repeated for the remaining grain types, as desired.


## Visualising total annual export to the United States, by grain type

```{r}
grains_annual %>%
  filter(str_detect(destination, pattern="United States")) %>%
  distinct(destination)
```

For the following, let us only consider the United States mainland (which I assume includes Alaska)
and extract the pertinent data.

```{r}
grains_annual_export_US <- grains_annual %>%
  filter(destination == "United States")
```

```{r}
ggplot(grains_annual_export_US, aes(x=Year, y=value, colour=type))+
  geom_line(alpha=0.6)
```


### Data preparation

The data preparation procedure is identical to the previous plots. We will need to:

- Scale the values on the y-axis to avoid scientific notation
- Reorder the grain types by their all-time maximums, in descending order

```{r}
grains_annual_export_US <- grains_annual_export_US %>%
  mutate(
    value = value / 1e6,
    type = fct_reorder(type, value, .fun=max, .desc=TRUE)
  )
```


### Building the plot

```{r, fig.height=9}
ggplot(grains_annual_export_US, aes(x=Year, y=value, colour=type))+
  geom_line(show.legend=FALSE)+
  gghighlight(use_direct_label=FALSE)+
  facet_wrap(~type)+
  labs(
    y="Amount exported (million tonnes)",
    title="Annual grain amounts exported to the United States, by grain type",
    subtitle="1966 to 2018 (where data is available)"
  )
```
