---
title: "Soft drink production"
output:
  html_document:
    anchor_sections: true
    highlight: pygments
    css: [css/base.css, css/downlit.css]
    md_extensions: -smart
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo=TRUE, warning=FALSE, message=FALSE,
  fig.align="center", fig.width=9.5, dev="svglite",
  class.source="sourceCode"
)

# Enable console highlighting, see `?fansi::set_knit_hooks`

old.hooks <- fansi::set_knit_hooks(knitr::knit_hooks)

styles <- c(
  getOption('fansi.style'),
  "PRE.fansi CODE {background-color: transparent;}",
  "PRE.fansi-error {background-color: transparent;}",
  "PRE.fansi-warning {background-color: transparent;}",
  "PRE.fansi-message {background-color: transparent;}"
)

old.hooks <- c(
  old.hooks,
  fansi::set_knit_hooks(
    knitr::knit_hooks,
    which=c('warning', 'error', 'message'),
    style=styles
  )
)

options(crayon.enabled = TRUE)
```

## Techniques employed

- Data wrangling:
  - Filtering data
  - Working with dates: months, quarters, years
  - Grouping and summarising data
  - Row-binding data

- Data visualisation:
  - Line plots, grouped line plots
  - Modifying the legend position and appearance


## About the data

From 1950 to 1977, soft-drink production was reported on a quarterly basis.

More information on this record can be found
[here](https://open.canada.ca/data/en/dataset/15dac000-240f-494c-86b6-311d8f8012b9).

The data and accompanying metadata can be downloaded together as a .zip file from
[here](https://www150.statcan.gc.ca/n1/tbl/csv/16100100-eng.zip).

From 1976 to 1995, soft-drink production was reported on a monthly basis.

More information on this record can be found
[here](https://open.canada.ca/data/en/dataset/de2e4ec3-dcb8-4dd1-aad8-3ccc1a61d633).

The data and accompanying metadata can be downloaded together as a .zip file from
[here](https://www150.statcan.gc.ca/n1/tbl/csv/16100099-eng.zip).


## Packages

```{r}
library(tidyverse)
library(lubridate)
theme_set(theme_bw())
```


## Obtaining the data

For these demos, I will be storing all data files in a folder called `data` with subfolders for
each topic.

```{r, eval=FALSE}
if (!dir.exists("./data/drinks")) {
  dir.create("./data/drinks", recursive=TRUE)
}

# Quarterly data from 1950 to 1977

download.file(
  "https://www150.statcan.gc.ca/n1/tbl/csv/16100100-eng.zip",
  destfile = "./data/drinks/qdrinks.zip"
)

unzip(
  "./data/drinks/qdrinks.zip",
  exdir = "./data/drinks"
)

# Monthly data from 1976 to 1995

download.file(
  "https://www150.statcan.gc.ca/n1/tbl/csv/16100099-eng.zip",
  destfile = "./data/drinks/mdrinks.zip"
)

unzip(
  "./data/drinks/mdrinks.zip",
  exdir = "./data/drinks"
)
```

```{r}
list.files("./data/drinks", pattern="csv")
```


## Reading in the quarterly data

```{r}
qdrinks <- read_csv("./data/drinks/16100100.csv")

glimpse(qdrinks)
```

The variables of interest are:

| Variable   | Description                         |
|------------|-------------------------------------|
| `REF_DATE` | Year and month of record            |
| `VALUE`    | Amount produced in thousand gallons |

As the dates are supplied as strings of format yyyy-mm, we can use `lubridate::ym()` to parse them
into proper dates.

```{r}
quarterly <- qdrinks %>%
  select(REF_DATE, VALUE) %>%
  rename(date = REF_DATE, quarterly_value = VALUE) %>%
  mutate(date = ym(date))

quarterly
```


## Reading in the monthly data

```{r}
mdrinks <- read_csv("./data/drinks/16100099.csv")

glimpse(mdrinks)
```

The variables of interest are:

| Variable   | Description                         |
|------------|-------------------------------------|
| `REF_DATE` | Year and month of record            |
| `VALUE`    | Amount produced in thousand gallons |

The monthly data can be processed similar to the quarterly data.

```{r}
monthly <- mdrinks %>%
  select(REF_DATE, VALUE) %>%
  rename(date = REF_DATE, monthly_value = VALUE) %>%
  mutate(date = ym(date))

monthly
```

**We want to expand the quarterly production data by converting our monthly data into quarterly
data, and appending it to the end of the original quarterly data. The final result should be a data
set that ranges from 1950 to 1995 with production values being reported each quarter.**


## Converting monthly data to quarterly data

Starting from the `monthly` data that we have, we create a new variable called `qdate` to keep track
of the quarter that each year-month combination belongs to. Grouping the data by the resulting
quarters and summing over the monthly production values results in the desired quarterly production
values. Finally, the quarters created under `qdate` are converted back to full dates in preparation
for the row binding of the two quarterly data sets.

```{r}
quarterly2 <- monthly %>%
  mutate(qdate = quarter(date, with_year=TRUE)) %>%
  group_by(qdate) %>%
  summarise(quarterly_value = sum(monthly_value)) %>%
  mutate(date = yq(qdate)) %>%
  select(date, quarterly_value)

quarterly2
```


## Combining the two quarterly data sets

Since there is an overlap in the dates between the quarterly and monthly data, we need to make sure
that all quarters are only included once when row binding `quarterly` and `quarterly2`. Let's find
the end date from the original quarterly data and keep only the observations from `quarterly2` that
occur after this date.

```{r}
quarterly_end_date <- quarterly %>%
  slice_max(date) %>%
  pull(date)

quarterly2 <- quarterly2 %>%
  filter(date > quarterly_end_date)
```

```{r}
full_quarterly <- bind_rows(quarterly, quarterly2)
```

### Checking our work

```{r}
full_quarterly %>%
  slice_min(date)
```

```{r}
full_quarterly %>%
  slice_max(date)
```

Since the data ranges from 1950 to 1995 (46 years), we should have 46 * 4 = 184 rows in our data.

```{r}
full_quarterly %>%
  nrow()
```

Looks good!


## Quarterly soft-drink production

In the original data, the production values were in the unit of thousands of gallons. Yet, the
values that we have are still quite large. Let's scale all the production values by one thousand
so that they will be in the units of million gallons.

```{r}
full_quarterly <- full_quarterly %>%
  mutate(quarterly_value = quarterly_value / 1e3)
```

```{r}
ggplot(full_quarterly, aes(x=date, y=quarterly_value))+
  geom_line()+
  labs(
    x="Date", y="Amount produced (million gallons)",
    title="Quarterly soft-drink production in Canada",
    subtitle="1950 to 1995"
  )
```

## Annual soft-drink production

An alternative to the previous plot is to colour the production values by quarter. In doing so,
the x-axis should be the year rather than the full date so that all the quarters of a given year
line up.

```{r}
full_quarterly %>%
  mutate(
    Quarter = as_factor(quarter(date)),
    Year = year(date)
  ) %>%
  ggplot(aes(x=Year, y=quarterly_value, colour=Quarter))+
  geom_line(alpha=0.6)+
  theme(legend.position="bottom")+
  guides(colour = guide_legend(label.position="bottom"))+
  labs(
    x="Year", y="Amount produced (million gallons)",
    title="Annual soft-drink production in Canada by quarter",
    subtitle="1950 to 1995"
  )
```

Finally, we can also look at the annual production without regard to the quarterly values.

```{r}
full_quarterly %>%
  mutate(Year = year(date)) %>%
  group_by(Year) %>%
  summarise(annual_value = sum(quarterly_value)) %>%
  ggplot(aes(x=Year, y=annual_value))+
  geom_line()+
  labs(
    x="Year", y="Amount produced (million gallons)",
    title="Annual soft-drink production in Canada",
    subtitle="1950 to 1995"
  )
```

---

#### Notes

In each usage of `group_by()` followed by `summarise()`, we never called `ungroup()`. This is
because `summarise()` automatically drops one grouping level after completion and in each usage, we
only grouped by a single variable. As such, after summarising, the data is fully ungrouped.
